#phydynR: Coalescent simulation and likelihood for phylodynamic inference


## Exponential growth example TODO 
Let's model a population which is growing exponentially with fixed per-capita birth rate `beta` and death rate `gamma`. 
Once we've defined the model, we can simulate trajectories and then simulate genealogies if sampling lineages at specific times. 
Finally, we'll see how to infer birth and/or death rates if the tree is observed (e.g. reconstructed from genetic sequence data). 

First, load the package and define the rates: 
```{r, echo=TRUE,results='hide'}
require(phydynR)
```
```{r}
births <- c(I = 'parms$beta * I' )
deaths <- c(I = 'parms$gamma * I' )
```
The rates are specified as a named vector of strings; names correspond to the names of the demes (`I` in this case). 
The strings are interpreted as R expressions, so should be written just as you would write other R code. 
The keywod `parms` that appears in the rate equations is a list of parameters, which can be set as we will see below. 

Now build the demographic process like so:
```{r}
dm <- build.demographic.process(births=births
  , deaths = deaths
  , parameterNames=c('beta', 'gamma') 
  , rcpp=FALSE
  , sde = TRUE)
```
Note the following

* You must specify the names of parameters that appear in `parms`. 
* The `sde` option allows you to speficfy if the model is stochastic (stochastic differential equations) or deterministic (ordinary differential equations)
* the `rcpp` option says if the equations are written as R code or as C code; we used R code in this case. 

Once the model is specified, we can simulate and visualise trajectories: 
```{r}
show.demographic.process( dm
 , theta = list( beta = 1.5, gamma = 1 )
 , x0  = c( I = 1 )
 , t0 = 0
 , t1 = 10 
) 
```
The variables `t0` and `t1` determine the time limits of integration. `x0` provides a named vector of initial condition; note well that names must correspond to those used to specify the rates. `theta` is a synonym for `parms` and provides the parameter list. 

You can also directly simulate the model using
```
 dm(theta, x0, t0, t1, res = 1000, integrationMethod = 'adams')
```
Here `integrationMethod` is based to the ODE solver and specifies which method to use (e.g. `euler` or `adams`). `res` is the number of time steps to use; larger values will be more accurate at expense of more computation. 

We can alternatively specify the equations as C code, in which case the model will be compiled using the `Rcpp` package. 
In this case, simulating the model will be very fast, but it will take a few seconds to compile the model. 
Let's compile a deterministic version of the model:
```{r}
dm.det <- build.demographic.process(births = c(I = 'beta * I')
  , deaths = c(I = 'gamma * I')
  , parameterNames=c('beta', 'gamma') 
  , rcpp=TRUE
  , sde = FALSE)
```
Note that when we use C code, the `parms` keyword is not used. 

Now let's simulate a coalescent tree conditioning on this demographic process: 
```{r}
tre <- sim.co.tree(   list( beta = 1.5, gamma = 1 )
  , dm.det
  , x0  = c(I = 1 )
  , t0 = 0
  , sampleTimes = seq(10, 15, length.out=50)
  , res = 1000
) 
```
This is self-explanotory except for the `sampleTimes` argument, which is required. 
This specifes the times relative to `t0` that each lineage is sampled. 
The length of this vector determines the sample size. 
This can be a named vector, in which case the taxon labels are retained in the returned tree. 
The returned tree is a `DatedTree` object which subclasses `ape::phylo`. So, most of the functions in the `ape` package will also work with the simulated tree. Let's plot it: 
```{r}
plot( ladderize( tre ))
```




## SIR example TODO 
## HIV example TODO 
