---
title: "Simulate genealogies using exponential growth and phydynR"
output:
  bookdown::html_vignette2:
  #bookdown::pdf_book:
    toc: TRUE
    fig_caption: TRUE
    citation_package: biblatex
    keep_tex: TRUE
    latex_engine: xelatex
bibliography: bib/tutorials.bib
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Simulate genealogies using exponential growth and phydynR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this tutorial we will use phydynR to simulate genealogies using the expontial
growth and to inder birth and/or death rate.

# Requirements

This tutorial assumes you have the following R packages already installed in
your computer.

* [phydynR](https://github.com/emvolz-phylodynamics/phydynR): implements the coalescent simulation and likelihood function for phylodynamic analyses.


## Load the R packages

```{r message = FALSE}
  library(phydynR)
  library(bbmle)

```

# The epidemiological model

Let's model a population which is growing exponentially with fixed per-capita birth rate `beta` and death rate `gamma`. 
Once we've defined the model, we can simulate trajectories and then simulate genealogies if sampling lineages at specific times. 
Finally, we will see how to infer birth and/or death rates if the tree is observed (e.g. reconstructed from genetic sequence data). 

First, let's define the rates: 

```{r}
# Define the birth rate
births <- c(I = 'parms$beta * I')
# Define the death rate
deaths <- c(I = 'parms$gamma * I')
```

The rates are specified as a named vector of strings; names correspond to the 
names of the demes (`I` in this case). The strings are interpreted as R expressions, 
so should be written just as you would write other R code. The keyword `parms` 
that appears in the rate equations is a list of parameters, which can be set as 
we will see below. 

Now build the demographic process like so:
```{r message = FALSE}
dm <- build.demographic.process(births = births,
                                deaths = deaths,
                                parameterNames=c('beta', 'gamma'),
                                rcpp = FALSE,
                                sde = TRUE)
```

Note the following:

* You must specify the names of parameters that appear in `parms`; 
* The `sde` option allows you to specify if the model is stochastic (stochastic differential equations) or deterministic (ordinary differential equations);
* the `rcpp` option says if the equations are written as R code or as C code; 
we used R code in this case. 

Once the model is specified, we can simulate and visualize trajectories: 

```{r fig.asp = 0.8, fig.width = 6, echo = TRUE}
show.demographic.process( dm,
                          theta = list( beta = 1.5, gamma = 1 ),
                          x0  = c( I = 1 ),
                          t0 = 0,
                          t1 = 10) 
```

The variables `t0` and `t1` determine the time limits of integration. 
`x0` provides a named vector of initial condition; note well that names must 
correspond to those used to specify the rates. `theta` is a synonym for `parms` 
and provides the parameter list. 

You can also directly simulate the model using

```{r eval = FALSE}
dm(theta = list( beta = 1.5, gamma = 1 ),
   x0 = c( I = 1) ,
   t0 = 0,
   t1 = 10,
   res = 1000, 
   integrationMethod = 'adams')
```

Here `integrationMethod` is based to the ODE solver and specifies which method to 
use (e.g. `euler` or `adams`). `res` is the number of time steps to use; larger 
values will be more accurate at expense of more computation. 

We can alternatively specify the equations as C code, in which case the model 
will be compiled using the `Rcpp` package. In this case, simulating the model will 
be very fast, but it will take a few seconds to compile the model. 
Let's compile a deterministic version of the model:

```{r}
dm.det <- build.demographic.process(births = c(I = 'beta * I'),
                                    deaths = c(I = 'gamma * I'),
                                    parameterNames=c('beta', 'gamma'),
                                    rcpp = TRUE,
                                    sde = FALSE)
```

Note that when we use C code, the `parms` keyword is not used. 

Now let's simulate a coalescent tree conditioning on this demographic process: 
```{r}
tre <- sim.co.tree(   list( beta = 1.5, gamma = 1 ),
                      dm.det,
                      x0  = c(I = 1 ),
                      t0 = 0,
                      sampleTimes = seq(10, 15, length.out = 50),
                      res = 1000) 
```

This is self-explanatory except for the `sampleTimes` argument, which is required. 
This specifies the times relative to `t0` that each lineage is sampled. 
The length of this vector determines the sample size. 
This can be a named vector, in which case the taxon labels are retained in 
the returned tree.  The returned tree is a `DatedTree` object which subclass 
`ape::phylo`. So, most of the functions in the `ape` package will also work 
with the simulated tree. Let's plot it: 

```{r fig.asp = 0.8, fig.width = 6, echo = TRUE}
plot(ladderize( tre ))
```

Finally, we will see how to compute a likelihood of parameters describing the 
demographic process given the tree as data. The `colik` function can be used to 
compute this likelihood, and model fitting can be done in a variety of ways. 
A Bayesian analysis could be done using the `mcmc` package, 
or maximum likelihood can be done using the `bbmle` package. 
Note that this returns the log likelihood. 

Here is an example invocation of the likelihood function using the simulated tree as data: 
```{r}
colik(tre,
      list( beta = 1.5, gamma = 1),
      dm.det,
      x0 = c( I = 1 ),
      t0 = -1,
      res = 1e3,
      timeOfOriginBoundaryCondition = FALSE)
```

Note the following options:

* `tre` is the `DatedTree` object that we fit to;
* `x0` is the initial conditions for the demographic process;
* `t0` is the time of origin of the process; you should choose a value that occurs 
before the root of the tree;
* `res` is the number of time steps used in the demographic process;
* `timeOfOriginBoundaryCondition` (default `FALSE`) : if `TRUE`, the function 
returns `-Inf` if the time of origin occurs before the root of the tree;
* `AgtYboundaryCondition` (default `TRUE`) : if `TRUE`, the function returns `-Inf` 
if the number of extant lineages in the tree is less than the simulated population 
size; this should usually be left as `TRUE`.

# Fitting the model

We will fit the model using a maximum likelihood approach.
We will focus on estimating the transmission rate `beta` and the initial size for
$I$. We will assume prior knowledge the death rate $\gamma$.

First, create the objective function to be minimized:

```{r echo = FALSE}
obj_fun <- function(beta, iniI){
 
  
  parms <- list(beta = beta, 
                I = iniI, 
                gamma = 1, 
                t0 = -1)
  
  
  mll <- -colik(tre,
                theta = parms,
                dm.det,
                x0 = c(I = 1),
                t0 = parms$t0,
                res = 1000)
  
  # track progress:
  print(c(mll, c(beta, I)))
  mll
}
```

We can then fit the model by running

```{r eval = FALSE}

fit <- mle2(obj_fun,
            start = list(beta = 2, iniI = 1),
            method='Nelder-Mead', 
            optimizer='optim',
            control = list(trace = 6, reltol = 1e-8))
```

Note the following options:

* `start` contains the starting conditions for the parameters we would like to 
estimate;
* `control` is a list of options for `optim`; see `?optim` for more information.

As this fit will take a little to run, we will load the results using

```{r echo = FALSE}
load( system.file('extdata/exponential_growth_fit.RData', package='phydynR') )
```

We can see the results by running

```{r}
AIC(fit)
logLik(fit)

#contains the parameter estimates
coef(fit)
```