---
title: "Estimating HIV transmission rates with phydynR"
output:
    rmarkdown::html_vignette:
    toc: TRUE
    fig_caption: true
    df_print: kable
    highlight: pygments
    citation_package: biblatex
    keep_tex: true
bibliography: bib/tutorials.bib
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{simulate_genelogies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette will demonstrate how to use structured coalescent models [@Volz2012] 
to estimate epidemiological parameters given a pathogen genealogy and discrete
trait information about sampled individuals. We will focus on the problem of
estimating HIV transmission rates by stage of infection and assuming we know
the stage of infection when patients are sampled and virus is sequenced.
We will fit a very simple model of an HIV epidemic using ordinary differential
equations (ODEs). 
In this model, the infectious period will be broken into three stages
of different average duration and with different transmission rates. The first
stage, early HIV infection (EHI) is short (average duration $1/\gamma_0 = 1$ year), 
but has high transmission rate $\beta_0$. The second stage, chronic HIV infection, 
is long (average duration $1/\gamma_1 = 7$ years), and has small transmission rate. 
The last stage, AIDS, lasts $1/\gamma_2 = 2$ years on average and has an intermediate
transmission rate. There are births into the susceptible state at rate $bN$ where
$N = S + I0 + I1 + I2$. And there is mortality due to natural causes from all states
at rate $\mu$. The parameter values are listed in table 1. The model equations are
as follows:


$\dot{S} = bN − \mu S − (\beta_0I_0 + \beta_1I_1 + \beta_2I_2)S/N$
$\dot{I_0} = (\beta_0I_0 + \beta_1I_1 + \beta_2I_2)S/N − (\mu + \gamma_0)I_0$
$\dot{I_1} = \gamma_0I_0 − (\mu + \gamma_1)I_1$
$\dot{I_2} = \gamma_1I_1 − (\mu + \gamma_2)I_2$

The model is also illustrated in figure 1.

The package is loaded by

```{r}
library(phydynR)
```


Create a list of the true parameter values:

```{r}
 parms_truth <- list(gamma0 = 1,
                     gamma1 = 1/7,
                     gamma2 = 1/2,
                     mu = 1/30,
                     b=.036,
                     beta0 = 12./10,
                     beta1 = 3./100,
                     beta2 = 9./100,
                     S_0 = 3000,
                     I0_0 = 1, 
                     I1_0 = 0.01, 
                     I2_0 = 0.01,
                     m = 3,
                     mm = 1)
```

In order to fit this model, we need to express the equations in a canonical
format. According to this format, we will tally birth and migration events
between demes. In our example, the deme corresponds to the stage of infection
that an infected host can be in, so we will refer the demes with the following
names:

```{r}
INFECTEDNAMES <- c('I0', 'I1', 'I2')
```

There are $m = 3$ demes in this model, so the birth events between demes
needs to be expressed with a $3 × 3$ matrix F. The element $F_kl$ represents the
rate of transmissions by a host in deme $k$ to a host in deme $l$. 
In our example, this is the following:

```{r}
births <- rbind(c('parms$beta0 * S * I0 / (S + I0 + I1 + I2)', '0', '0'),
                c('parms$beta1 * S * I1 / (S + I0 + I1 + I2)', '0', '0'),
                c('parms$beta2 * S * I2 / (S + I0 + I1 + I2)', '0', '0'))

rownames(births)=colnames(births)<- INFECTEDNAMES
```

Each element of the matrix is a string that will be parsed as R code and evaluated, 
so it is important to write it exactly as you would if you were solving
the equations outside of phydynR. Also note that the object parms is accessible
to these equations, which is a list of parameters - this may include parameters
to be estimated. Note that there are zero rates in the 2nd and third columns,
since all new infected hosts start out in the first stage of infection (EHI). 
Also note that we must give row and column names to the matrix, and these names
must correspond to the names of the demes. 

Similarly, we must create a matrix of migrations:

```{r}
 migrations <- rbind(c('0', 'parms$gamma0 * I0', '0'),
                     c('0', '0', 'parms$gamma1 * I1'),
                     c('0', '0', '0'))

rownames(migrations) = colnames(migrations) <- INFECTEDNAMES
```

Note that this matrix tallys the stage progression from EHI to chronic and from
chronic to AIDS.
We must also write a vector of expressions for events that terminate a
lineage – In this model, this occurs due to natural or AIDS related mortality:

```{r}
deaths <- c('parms$mu*I0',
            'parms$mu*I1',
            'parms$mu*I2 + parms$gamma2 * I2')

names(deaths) <- INFECTEDNAMES
```

Finally, we must write a vector of ODEs for state variables that do not
correspond to demes in the coalescent model. In our example, there is only one
such variable- the number of susceptibles:

```{r}
nonDemeDynamics <- paste(sep = '',
                         '-parms$mu*S + parms$mu*(S + I0 + I1 + I2)',
                         '- S * (parms$beta0*I0+parms$beta1*I1+parms$beta2*I2) / (S + I0 + I1 + I2)')

names(nonDemeDynamics) <- 'S'
```

Note well that in all cases, the expression or equation must have the 
corresponding name of the state variable.
The model can be fitted to a binary tree with dated tips. We will use a
simulated tree such that we know the true parameter values and population size
through time. We load this by:

```{r}
# read the tree
tree <- read.tree(system.file('extdata/hivSimulation.nwk', package='phydynR'))
#~ the sample times are the same, because it is a homochronous sample at 50 years
sampleTimes <- rep(50, length(tree$tip.label))
names(sampleTimes) <- tree$tip.label
# create a tree with dated tips and internal nodes,
# will infer the sample states from tip labels
bdt <- DatedTree(phylo = tree, 
                 sampleTimes = sampleTimes)
bdt
```