---
title: "Estimating HIV transmission rates with phydynR"
author: "Erik Volz"
date: "`r Sys.Date()`"
output:
  bookdown::html_vignette2:
    toc: TRUE
    fig_caption: true
    df_print: kable
    highlight: pygments
    citation_package: biblatex
    keep_tex: true
bibliography: bib/tutorials.bib
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{transmission_rates}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "latex")
```

# Introduction

This vignette will demonstrate how to use structured coalescent models [@Volz2012] 
to estimate epidemiological parameters given a pathogen genealogy and discrete
trait information about sampled individuals.

# Requirements

This tutorial assumes you have the following R packages already installed in
your computer.

* [phydynR](https://github.com/emvolz-phylodynamics/phydynR): implements the coalescent simulation and likelihood function for phylodynamic analyses
* [bbmle](https://github.com/bbolker/bbmle): tools for general likelihood estimation

## Load the R packages

```{r message = FALSE}
  library(phydynR)
  library(bbmle)
```

# The epidemiological model

We will focus on the problem of estimating HIV transmission rates by stage of 
infection and assuming we know the stage of infection when patients are sampled 
and virus is sequenced. We will fit a very simple model of an HIV epidemic using 
ordinary differential equations (ODEs).
In this model, the infectious period will be broken into three stages
of different average duration and with different transmission rates. The first
stage, early HIV infection (EHI) is short (average duration $1/\gamma_0 = 1$ year), 
but has high transmission rate $\beta_0$. The second stage, chronic HIV infection, 
is long (average duration $1/\gamma_1 = 7$ years), and has small transmission rate. 
The last stage, AIDS, lasts $1/\gamma_2 = 2$ years on average and has an intermediate
transmission rate. There are births into the susceptible state at rate $bN$ where
$N = S + I_0 + I_1 + I_2$. And there is mortality due to natural causes from all states
at rate $\mu$. The parameter values are listed in table \@ref(tab:table1). The model equations are:


$\dot{S} = bN − \mu S − (\beta_0I_0 + \beta_1I_1 + \beta_2I_2)S/N$

$\dot{I_0} = (\beta_0I_0 + \beta_1I_1 + \beta_2I_2)S/N − (\mu + \gamma_0)I_0$

$\dot{I_1} = \gamma_0I_0 − (\mu + \gamma_1)I_1$

$\dot{I_2} = \gamma_1I_1 − (\mu + \gamma_2)I_2$


```{r table1, echo = FALSE}
Parameter = c("Duration EHI",
              "Duration chronic",
              "Duration AIDS",
              "Birth rate",
              "Natural death rate",
              "EHI transmission rate",
              "Chronic transmission rate",
              "AIDS transmission rate",
              "Initial number of susceptibles") 
Symbol = c("$1/\\gamma_0$",
           "$1/\\gamma_1$",
           "$1/\\gamma_2$",
           "$b$",
           "$\\mu$",
           "$\\beta_0$",
           "$\\beta_1$",
           "$\\beta_2$",
           "$S(0)$") 
Values = c("1 year",
           "7 years",
           "2 years",
           "0.036",
           "$1/30$",
           "1.2",
           "0.03",
           "0.09",
           "3000") 
data_list = data.frame(Parameter, Symbol, Values) 


kbl(data_list, booktabs = TRUE, escape = FALSE, 
    caption = "Parameter symbols and values")

```


Create a list of the true parameter values:



```{r}
 parms_truth <- list(gamma0 = 1,
                     gamma1 = 1/7,
                     gamma2 = 1/2,
                     mu = 1/30,
                     b = 0.036,
                     beta0 = 12./10,
                     beta1 = 3./100,
                     beta2 = 9./100,
                     S_0 = 3000,
                     I0_0 = 1, 
                     I1_0 = 0.01, 
                     I2_0 = 0.01,
                     m = 3,
                     mm = 1)
```

In order to fit this model, we need to express the equations in a canonical
format. According to this format, we will tally birth and migration events
between demes. In our example, the deme corresponds to the stage of infection
that an infected host can be in, so we will refer the demes with the following
names:

```{r}
INFECTEDNAMES <- c('I0', 'I1', 'I2')
```

There are $m = 3$ demes in this model, so the birth events between demes
needs to be expressed with a $3 × 3$ matrix F. The element $F_kl$ represents the
rate of transmissions by a host in deme $k$ to a host in deme $l$. 
In our example, this is the following:

```{r}
births <- rbind(c('parms$beta0 * S * I0 / (S + I0 + I1 + I2)', '0', '0'),
                c('parms$beta1 * S * I1 / (S + I0 + I1 + I2)', '0', '0'),
                c('parms$beta2 * S * I2 / (S + I0 + I1 + I2)', '0', '0'))

rownames(births) = colnames(births) <- INFECTEDNAMES
```

Each element of the matrix is a string that will be parsed as R code and evaluated, 
so it is important to write it exactly as you would if you were solving
the equations outside of phydynR. Also note that the object parms is accessible
to these equations, which is a list of parameters - this may include parameters
to be estimated. Note that there are zero rates in the 2nd and third columns,
since all new infected hosts start out in the first stage of infection (EHI). 
Also note that we must give row and column names to the matrix, and these names
must correspond to the names of the demes. 


Similarly, we must create a matrix of migrations:

```{r}
migrations <- rbind(c('0', 'parms$gamma0 * I0', '0'),
                     c('0', '0', 'parms$gamma1 * I1'),
                     c('0', '0', '0'))

rownames(migrations) = colnames(migrations) <- INFECTEDNAMES
```

Note that this matrix tallys the stage progression from EHI to chronic and from
chronic to AIDS.

We must also write a vector of expressions for events that terminate a
lineage – In this model, this occurs due to natural or AIDS related mortality:

```{r}
deaths <- c('parms$mu*I0',
            'parms$mu*I1',
            'parms$mu*I2 + parms$gamma2 * I2')

names(deaths) <- INFECTEDNAMES
```

Finally, we must write a vector of ODEs for state variables that do not
correspond to demes in the coalescent model. In our example, there is only one
such variable- the number of susceptibles:

```{r}
nonDemeDynamics <- paste(sep = '',
                         '-parms$mu*S + parms$mu*(S + I0 + I1 + I2)',
                         '- S * (parms$beta0*I0+parms$beta1*I1+parms$beta2*I2) / (S + I0 + I1 + I2)')

names(nonDemeDynamics) <- 'S'
```

Note well that in all cases, the expression or equation must have the 
corresponding name of the state variable.
The model can be fitted to a binary tree with dated tips. We will use a
simulated tree such that we know the true parameter values and population size
through time. We load this by:

```{r echo = FALSE}
# read the tree
tree <- read.tree(system.file('extdata/hivSimulation.nwk', package='phydynR'))

# the sample times are the same, because it is a homochronous sample at 50 years
sampleTimes <- rep(50, length(tree$tip.label))
names(sampleTimes) <- tree$tip.label

# create a tree with dated tips and internal nodes,
# will infer the sample states from tip labels
bdt <- DatedTree(phylo = tree, 
                 sampleTimes = sampleTimes,
                 minEdgeLength = 0.01)
bdt
```

Note well that the vector of sample times must have the names of each taxon. 
If the state of each taxon (the stage of infection of each sample unit) is encoded 
at the end of taxon name, it can be loaded automatically by passing the argument 
_sampleStatesAnnotations_. Alternatively, that information can be passed as a
matrix with row names corresponding to the tip labels of the tree.

After setting up all the components of the mathematical model, we can build the 
demographic process using the function `build.demographic.process` from the `phydynR` 
package. The `dm` output can be used as input to coalescent models for the 
calculation of the likelihood.

```{r}
dm <- build.demographic.process(births = births,
                                nonDemeDynamics = nonDemeDynamics,
                                migrations = migrations,
                                deaths = deaths,
                                parameterNames = names(parms_truth),
                                rcpp = FALSE,
                                sde = FALSE)
```

Now we can calculate the likelihood of the tree and see how long it takes
by using the `system.time` function:

```{r eval = FALSE}
print(system.time(print(
  colik(bdt,
        theta = parms_truth,
        dm,
        x0=c(I0 = 1, I1 = 0.01, I2 = 0.01, S = parms_truth$S_0),
        t0 = 0,
        res = 1000,
        integrationMethod = 'euler')
  )))
```

It is a bit slow, so note that there is an alternative low-level interface that 
can be a bit faster (`colik.pik.fgy`). Also note that changing
the _integrationMethod_ (choose ‘euler’), _maxHeight_ (only fit to part of
the tree) and _res_ (set to a smaller value) options can dramatically
speed up the calculation at the cost of some accuracy.
We can fit the model by using the `bbmle` or `stats4` package.

We will focus on estimating the three transmission rates of the system along with
a nuisance parameter that controls initial conditions, $t_0$, which is the time of
origin of the epidemic. We will assume prior knowledge of the stage progression
rates $\gamma_i$ mortality rates $\mu$, and susceptible population size $S(0)$.

First, create the objective function to be minimized:

```{r}
obj_fun <- function(lnbeta0, lnbeta1, lnbeta2, t0){
  
  parms <- parms_truth
  parms$beta0 <- exp(lnbeta0)
  parms$beta1 <- exp(lnbeta1)
  parms$beta2 <- exp(lnbeta2)
  
  mll <- -colik(bdt,
                theta = parms,
                dm,
                x0=c(I0 = 1, I1 = 0.01, I2 = 0.01, S = parms_truth$S_0),
                t0 = 0,
                res = 1000,
                integrationMethod = 'rk4')

    # track progress:
  print(c(mll, exp(c(lnbeta0, lnbeta1, lnbeta2) ), t0) )
  mll
}
```

Note that this uses log-transformation for variables that must be positive (like
rates).

We can then fit the model by running

```{r eval = FALSE}
fit <- mle2(obj_fun,
            start = list(lnbeta0 = log(0.6), 
                         lnbeta1 = log(0.2), 
                         lnbeta2 = log(0.05), 
                         t0 = 0),
            method='Nelder-Mead', 
            optimizer='optim',
            control = list(trace = 6, reltol = 1e-8))
```

Note that we are starting the optimizer far from the true parameter values. If
fitting a model to real data, it is recommended to try many different starting
conditions over a large range of values. The optimizer would take about 10
minutes to run, so instead we will load the results:

```{r}
load( system.file('extdata/hivModel0-fit.RData', package='phydynR') )

AIC(fit)
logLik(fit)
coef(fit)
exp(coef(fit))
```

We can compare the fitted model to the true number of infected through
time, which is shown in \@ref(fig:figure1)

```{r figure1, fig.cap = "The actual (points) and estimated (lines) number of infections through time. Black: EHI, Red: chronic, Green: AIDS", echo = TRUE}
parms <- parms_truth
parms[c('beta0', 'beta1', 'beta2', 't0')] <- unname(exp(coef(fit)))
parms$t0 <- log(parms$t0)
ox <- solve.model(t0 = coef(fit)['t0'], 
                  t1 = bdt$maxSampleTime, 
                  x0 = c(I0 = 1, I1 =0.01, I2 = 0.01, S = parms_truth$S_0), 
                  births,  
                  deaths, 
                  nonDemeDynamics, 
                  parms, migrations=migrations, integrationMethod = 'rk4')
matplot( rtimes, rY, xlab='Time', ylab='Number infected by stage')
lines(ox[,1] , ox[, 'I0'], col='black')
lines(ox[,1] , ox[, 'I1'], col='red')
lines(ox[,1] , ox[, 'I2'], col='green')
```


We can calculate a confidence interval for the transmission rates using likelihood 
profiles:

```{r eval = FALSE}
profbeta <- profile(fit, 
                    which = 1, 
                    alpha = 0.05, 
                    std.err = 0.5, 
                    trace = TRUE, 
                    tol.newmin = 1 )
```

This takes a long time, so we will load the results:

```{r}
load( system.file('extdata/hivModel0-profbeta.RData', package='phydynR') )
```

We see that the confidence interval covers the true value:

```{r}
 c(exp(confint(profbeta )), TrueVal = parms_truth$beta0)
```

And, we can visualize the profile (\@ref(fig:figure2)).

```{r eval = FALSE}
plot(profbeta)
abline( v = log( parms_truth$beta0) , col='red')
```



```{r out.height = "460px", fig.align = "center", fig.cap = "Likelihood profile for the transmission rate $\\beta_0$ with confidence levels. The true parameter value is indicated by the vertical red line.", echo = FALSE}

knitr::include_graphics("figures/sir_vignette-proffigure.pdf")
```

