---
title: "Estimating transmission rate using the SIR model and phydynR"
author: "Erik Volz"
date: "`r Sys.Date()`"
output:
  bookdown::html_vignette2:
  #bookdown::pdf_book:
    toc: TRUE
    fig_caption: TRUE
    citation_package: biblatex
    keep_tex: TRUE
    latex_engine: xelatextex
bibliography: bib/tutorials.bib
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Estimating transmission rate using the SIR model and phydynR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(kableExtra)
library(magrittr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette will demonstrate how to use coalescent models as described
in [@Volz2012] to estimate transmission rate parameters given a pathogen genealogy.

# Basic requirements
This vignette assumes that you have the following packages already installed:

* ape: R package to work with phylogenetic trees.
* rjason: Converts R objects into JSON objects and vice-versa.
* [phydynR]("https://github.com/emvolz-phylodynamics/phydynR"): implements the coalescent simulation and likelihood function for phylodynamics analysis

## Load the necessary packages:
```{r message = FALSE}
  library(phydynR)
  library(rjson)
```


Suppose an epidemic occurs according to a density-dependent susceptible-
infected-recovered (SIR) process, and a given infected individual generates a 
new infection at the rate $\beta SI$, where $S$ is the number susceptible and $\beta$
is the transmission rate. Furthermore, infected individuals will be removed from 
the population at per capita rate $\gamma$. At a single point in time, 
a random sample of $n = 75$ infected individuals is taken and the genealogy is 
reconstructed from the history of transmissions. We have simulated such a dataset 
using MASTER 1.10 [@Vaughan2013], which can be loaded by


```{r}
tree <- read.tree(system.file("extdata/sirModel0.nwk", package = "phydynR"))
```

And, the epidemic trajectory information can be loaded by

```{r}
epidata <- fromJSON(file=system.file("extdata/sirModel0.json", package = "phydynR"))
```

The true parameter values are given in table 1. The file used to simulate the data
can be viewed by 

```{r}
file.show( system.file("extdata/sirModel0.xml", package = "phydynR")) 
```

```{r table1, echo = FALSE}
Parameter = c("Duration infection",
              "Transmission rate",
              "Population size",
              "Initial num infected",
              "Time of sampling") 
Symbol = c("$1/\\gamma$",
           "$\\beta$",
           "$S(0)$",
           "$I(0)$",
           "$T$") 
Values = c("1",
           "2.0002e-4",
           "9999",
           "1",
           "12") 
data_list = data.frame(Parameter, Symbol, Values) 


kbl(data_list, booktabs = TRUE, escape = FALSE, 
    caption = "Parameter symbols and values")

```


We will fit a simple ordinary differential equation (ODE) model to the genealogy:

$\dot{S} = -\beta SI$
$\dot{I} = \beta SI -\gamma I$

Relevant parameters of the system are the transmission rate $\beta$, 
recovery rate $\gamma$, initial population size $S(0)$ and initial number 
infected $I(0)$. Not all parameters are identifiable from these data, so we will
assume prior knowledge of $S(0)$ and $\gamma$ and focus on estimating $\beta$
and the nuisance parameter $I(0)$. Note that an imprecise estimate of $S(0)$
is also possible.

Create a list to store the true parameter values:

```{r}

parms_truth <- list( beta = 0.00020002, 
                     gamma = 1, 
                     S0 = 9999, 
                     t0 = 0 )

```

Note that the true value of R0 is $\beta S(0)/\gamma = 2$.

And, create a tree with dated tips and internal nodes:

```{r}

sampleTimes <- rep(12, 75)
names(sampleTimes) <- tree$tip.label

bdt <- DatedTree( phylo = tree, 
                  sampleTimes = sampleTimes)
bdt
```

Note that the vector of sample times must have names corresponding to the
taxon labels in tree.

In order to fit this model, we need to express the equations in a canonical
format:

```{r}
births <- c( I = "beta * S * I" )
deaths <- c( I = "gamma * I" )
nonDemeDynamics <- c(S = "-beta * S * I")
```

The births vector gives the total rate that all infected generate new infections
and deaths gives the rate that lineages are terminated. 
The nonDemeDynamics vector gives the equations for state variables that are not
directly involved in the genealogy (e.g. because a pathogen never occupies a 
susceptible host by definition).
Each element of the vectors is a string that will be parsed as R code and 
evaluated, so it is important to write it exactly as you would if you were 
solving the equations in R. 
Also note that the object parms is accessible to these equations, which is a 
list of parameters - this may include parameters to be estimated. Also
note that we must give names to the vectors, and these names must correspond
to the names of the demes.

We will use these initial conditions

```{r}
# initial number of I and S
x0 <- c(I = 1, S = unname(parms_truth$S0))

# initial t0 (time of origin of the process)
t0 <- bdt$maxSampleTime - max(bdt$heights) - 1
```

The time of origin t0 is chosen somewhat arbitrarily, but should occur before
the root of the tree.


## The demographic model

After setting up all the components of the mathematical model, we can build the 
demographic process using the function `build.demographic.process` from the `phydynR` 
package. The `dm` output can be used as input to coalescent models for the 
calculation of the likelihood.

```{r}
dm <- build.demographic.process(births = births,
                                nonDemeDynamics = nonDemeDynamics,
                                deaths = deaths,
                                parameterNames = names(parms_truth),
                                rcpp = TRUE,
                                sde = FALSE)
```


Now we can calculate the likelihood of the tree and assess how long it takes:

```{r}
print(system.time(print(phydynR::colik(tree = bdt,
                                       theta = parms_truth,
                                       demographic.process.model = dm,
                                       x0 = x0,
                                       t0 = t0,
                                       res = 1000,
                                       integrationMethod = "rk4")
                        )))
```


Note that changing the integrationMethod (choose "euler"), maxHeight
(only fitt to part of the tree) and res (set to a smaller value) options
can dramatically speed up the calculation at the cost of some accuracy.

We can fit the model using maximum likelihood with the bbmle or stats4
R packages.

```{r}
obj_fun <- function(lnbeta, lnI0){
  
  beta <- exp(lnbeta)
  I0 <- exp(lnI0)
  parms <- parms_truth
  parms$beta <- beta
  x0 <- list(I=unname(I0), S = unname(parms$S0) )
  
  mll <- -phydynR::colik(tree = bdt,
                         theta = parms_truth,
                         demographic.process.model = dm,
                         x0 = x0,
                         t0 = t0,
                         res = 1000,
                         integrationMethod = "rk4")
  
  print(paste(mll, beta, I0))
  mll
}
```


